Descriptive Complexity for Counting Complexity Classes

### 2017-03-29 LMCS/proof-capture-fpspace.tex
journal 29 mar

For condition (1), notice that each $\pfp$ formula can be evaluated in deterministic polynomial space. Also, the constant function $f = s$ can be trivially simulated in $\fpspace$ for each $s\in\nat$. Furthermore, $\fpspace$ is closed under exponential sum and multiplication. This suffices to prove that the condition holds.

>>>>

For condition (1), notice that each $\pfp$ formula can be evaluated in deterministic polynomial space, that the constant function $s$ can be trivially simulated in $\fpspace$ for each $s\in\nat$, and that $\fpspace$ is closed under exponential sum and multiplication. This suffices to show that the condition holds.

### 2017-03-29 LMCS/proof-capture-optp.tex
journal 29 mar

Similar than the previous proofs, we construct recursively a non-deterministic polynomial time Turing machine $M_{\alpha}$ with output tape for each $\maxqso(\fo)$ formula $\alpha$ over a signature $\R$.
This machine, on input $(\A,v,V)$, non-deterministically produces $\sem{\alpha}(\A,v,V)$ (in binary) over the output tape of some run, and this value is the maximum value over all runs.
Suppose $\A$ has domain $A$. If $\alpha$ is a $\fo$-formula $\varphi$, then the machine checks if $(\A,v,V)\models\varphi$ in deterministic polynomial time, and output $1$ if, and only if, $(\A,v,V)$ satisfies $\varphi$. If $\alpha$ is a constant $s$, it outputs $s$ in binary over the output tape. If $\alpha = (\beta \add \gamma)$, then it simulates $M_{\beta}$ and $M_{\gamma}$ on input $(\A,v,V)$, adds the output of both machines and prints this value over the output tape. If $\alpha = (\beta \mult \gamma)$ or $\alpha = \max\{\beta,\gamma\}$, then it does the same than the previous case but it multiplies or maximizes, respectively, the outputs of both machines instead of adding. 
If $\alpha = \sa{x}\beta$, it iterates over all elements $a\in A$ simulating and adding the output of $M_{\beta}$ on input $(\A,v[a/x],V)$. $M_{\alpha}$ finally outputs the aggregated value in the output tape. 
If $\alpha = \pa{x}\beta$ or $\alpha = \maxa{x}\beta$, then $M_{\alpha}$ does the same idea than the previous case with the difference that the output of $M_{\beta}$ on input $(\A,v[a/x],V)$ is multiplied or maximized, respectively.
If $\alpha = \maxa{X}\beta$, it chooses $B\in A^{arity(X)}$ and simulates $M_{\beta}$ on input $(\A,v,V[B/X])$.
This covers all possible cases for $\alpha$.
Furthermore, it is straightforward to prove that each of these steps are correct and can be computed with a non-deterministic polynomial time Turing machine with output tape.
Let $\alpha$ be a formula in $\maxqso$ over a signature $\R$ and let $f$ be a function over $\R$ such that $f(\enc(\A))$ is equal to the maximum run (with respect to the output value) of $M_{\alpha}$ on input $(\A,v,V)$ for some $(\A,v,V) \in \ostr[\R]^*$. We have that $f$ is a $\maxp$ function over $\R$ and that $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A\in\ostr[\R]$.
Finally, one can easily see that the same construction holds with $\minqso(\fo)$ by constructing a Turing machine that take the min over all runs instead of max.

>>>>

In this proof we construct recursively a non-deterministic polynomial time Turing machine $M_{\alpha}$ with output tape for each $\maxqso(\fo)$ formula $\alpha$ over a signature $\R$.
This machine, on input $(\A,v,V)$, non-deterministically produces $\sem{\alpha}(\A,v,V)$ (in binary) over the output tape of some run, and this value is the maximum value over all runs.
Suppose $\A$ has domain $A$. If $\alpha$ is a $\fo$-formula $\varphi$, then the machine checks if $(\A,v,V)\models\varphi$ in deterministic polynomial time, and output $1$ if, and only if, $(\A,v,V)$ satisfies $\varphi$. If $\alpha$ is a constant $s$, it outputs $s$ in binary over the output tape. If $\alpha = (\beta \add \gamma)$, then it simulates $M_{\beta}$ and $M_{\gamma}$ on input $(\A,v,V)$, adds the output of both machines and prints this value over the output tape. If $\alpha = (\beta \mult \gamma)$ or $\alpha = \max\{\beta,\gamma\}$, then it does the same than the previous case but it multiplies or maximizes, respectively, the outputs of both machines instead of adding. 
If $\alpha = \sa{x}\beta$, it iterates over all elements $a\in A$ simulating and adding the output of $M_{\beta}$ on input $(\A,v[a/x],V)$. $M_{\alpha}$ finally outputs the aggregated value in the output tape. 
If $\alpha = \pa{x}\beta$ or $\alpha = \maxa{x}\beta$, then $M_{\alpha}$ does the same idea than the previous case with the difference that the output of $M_{\beta}$ on input $(\A,v[a/x],V)$ is multiplied or maximized, respectively.
If $\alpha = \maxa{X}\beta$, it chooses $B\in A^{arity(X)}$ and simulates $M_{\beta}$ on input $(\A,v,V[B/X])$.
This covers all possible cases for $\alpha$.
Furthermore, it is straightforward to prove that each of these steps are correct and can be computed with a non-deterministic polynomial time Turing machine with output tape.
Let $\alpha$ be a formula in $\maxqso$ over a signature $\R$ and let $f$ be a function over $\R$ such that $f(\enc(\A))$ is equal to the maximum run (with respect to the output value) of $M_{\alpha}$ on input $(\A,v,V)$ for some $(\A,v,V) \in \ostr[\R]^*$. We have that $f$ is a $\maxp$ function over $\R$ and that $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A\in\ostr[\R]$.
Finally, one can easily see that the same construction holds with $\minqso(\fo)$ by constructing a Turing machine that take the min over all runs instead of max.

### 2017-03-29 LMCS/proof-capture-spanP.tex
journal 29 mar

To prove the condition (2), we use the fact that $\spp = \#\eso$. The condition holds using the same argument as in proposition \ref{prop:capture-shP}.

For condition (1), notice that given an $\eso$ formula $\varphi$, checking whether $\A\models\varphi$ can be done in non-deterministic polynomial time on the size of $\A$\cite{fagin1974generalized}. To ensure that this $\spp$ machine adds at most one different output, the same string is written as output in each non-deterministic path. Furthermore, the constant function $f = s$ for some $s\in\nat$ can be trivially simulated in $\spp$. The condition holds analogously to proposition \ref{prop:capture-shP} since $\spp$ possesses the mentioned closure properties as well\cite{OH93}.

>>>>

To prove the condition (2), we use the fact that $\spp = \#(\eso)$. The condition holds using the same argument as in proposition \ref{prop:capture-shP}.

For condition (1), notice that given an $\eso$ formula $\varphi$, checking whether $\A\models\varphi$ can be done in non-deterministic polynomial time on the size of $\A$\cite{fagin1974generalized}. The $\spp$ machine we construct to evaluate the formula produces at most one different output, and does this by writing the same string as output in each non-deterministic path. Furthermore, the constant function $s$ for some $s\in\nat$ can be trivially simulated in $\spp$. The condition holds analogously to proposition \ref{prop:capture-shP} since $\spp$ possesses the mentioned closure properties as well\cite{OH93}.

### 2017-03-29 LMCS/proof-theo-pi1-pnf.tex
journam 7 mar
journal 29 mar

\vspace{1em}
First, we prove that the formula $\alpha_{0} = \left( \sa{X} 1 \right) + 1$ with $\arity(X) = 1$ (i.e. the function $2^{n}+1$, where $n$ is the size of the input structure) is not equivalent to any formula in $\loge{0}$-PNF. Suppose that there exists some formula $\alpha = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ in $\loge{0}$-PNF that is equivalent to the $\eqso(\loge{0})$ formula $\alpha_0$.
In \cite{SalujaST95}, it was proved that if $\length{\bar{X}} > 0$, then the function defined by $\alpha$, for big enough structures, is always even which is not possible in our case.
On the other hand, if $\alpha$ is of the form $\sa{\bar{x}}\varphi(\bar{x})$, then $\alpha$ defines a polynomially bounded function which also leads to a contradiction.

\vspace{1em}
Second, we prove that the formula $\alpha_{1} = 2$ (i.e. the constant $2$) is not equivalent to any formula in $\loge{1}$-PNF. Suppose that there exists some formula $\alpha = \sa{\bar{X}}\sa{\bar{x}}\exists\bar{y}\, \varphi(\bar{X},\bar{x},\bar{y})$ in $\loge{1}$-PNF that is equivalent to the $\eqso(\loge{1})$ formula $2$. 
First, if $\length{\bar{X}} = \length{\bar{x}} = 0$, then the function defined by $\alpha$ is never greater than 1. 
Therefore, suppose that $\length{\bar{X}} > 0$ or $\length{\bar{x}} > 0$, and consider any ordered structure $\A$. 
Since $\sem{\alpha}(\A) = 2$, there exist at least two assignments $(\bar{B}_1,\bar{b}_1,\bar{a}_1)$, $(\bar{B}_2,\bar{b}_2,\bar{a}_2)$ to $(\bar{X},\bar{x},\bar{y})$ such that for both, $\A\models\varphi(\bar{B}_i,\bar{b}_i,\bar{a}_i)$. Now consider the ordered structure $\A'$ that is obtained by duplicating $\A$. This is, each half of $\A'$ is isomorphic to $\A$. Note that $\A'\models\varphi(\bar{B}_i,\bar{b}_i,\bar{a}_i)$ for $i = 1,2$ and there exists a third assignments $(\bar{B}_1',\bar{b}_1',\bar{a}_1')$ that is isomorphic to $(\bar{B}_1,\bar{b}_1,\bar{a}_1)$ but in the other half to the structure such that $\A'\models\varphi(\bar{B}_1',\bar{b}_1',\bar{a}_1')$. We have that $\sem{\alpha}(\A) \geq 3$ which is a contradiction.

@@

\sa{\bar{X}}\sa{\bar{x}}\sa{\bar{Y}}\sa{\bar{y}}[&(\varphi(\bar{X},\bar{x})\wedge\exists\bar{v}(\bar{v}_0 < \bar{x} \vee \bigvee_{i = 1}^{\length{\bar{X}}}X(\bar{v}_i))\wedge\bigwedge_{i = 1}^{\length{\bar{Y}}} \forall\bar{z}\neg Y_i(\bar{z}) \wedge \forall\bar{z}(\bar{y}\leq\bar{z})\vee\\

@@

\sa{\bar{X}}\sa{\bar{Y}}\sa{\bar{x}}\sa{\bar{y}}\sa{\bar{u}}[&(\varphi(\bar{X},\bar{x})\wedge(\bar{v}_0 < \bar{x} \vee \bigvee_{i = 1}^{\length{\bar{X}}}X(\bar{v}_i))\wedge \forall\bar{u}'((\bar{v}_0' < \bar{x} \vee \bigvee_{i = 1}^{\length{\bar{X}}}X(\bar{v}_i'))\to\bar{u}\leq\bar{u}') \wedge \\
& \hspace{3cm}  \bigwedge_{i = 1}^{\length{\bar{Y}}} \forall\bar{z}\neg Y_i(\bar{z}) \wedge \forall\bar{z}(\bar{y}\leq\bar{z})\vee\\

@@

& \hspace{3cm} (\bar{y} < \bar{w}_0 \vee \bigvee_{i = 1}^{\length{\bar{Y}}}\neg Y(\bar{w}_i))\wedge\forall\bar{u}'(\bar{y} < \bar{w}_0' \vee \bigvee_{i = 1}^{\length{\bar{Y}}}\neg Y(\bar{w}_i'))\to\bar{u}\leq\bar{u}')\vee \\

>>>>

\vspace{1em}
First, we prove that the formula $\alpha_{0} = \left( \sa{X} 1 \right) + 1$ with $\arity(X) = 1$ (i.e. the function $2^{\vert\A\vert}+1$) is not equivalent to any formula in $\loge{0}$-PNF. Suppose that there exists some formula $\alpha = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ in $\loge{0}$-PNF that is equivalent to $\alpha_0$.
In \cite{SalujaST95}, it was proved that if $\length{\bar{X}} > 0$, the function defined by $\alpha$ is always even for big enough structures, which is not possible in our case.
On the other hand, if $\alpha$ is of the form $\sa{\bar{x}}\varphi(\bar{x})$, then $\alpha$ defines a polynomially bounded function which also leads to a contradiction.

\vspace{1em}
Second, we prove that the formula $\alpha_{1} = 2$ (i.e. the constant function $2$) is not equivalent to any formula in $\loge{1}$-PNF. Suppose that there exists some formula $\alpha = \sa{\bar{X}}\sa{\bar{x}}\exists\bar{y}\, \varphi(\bar{X},\bar{x},\bar{y})$ in $\loge{1}$-PNF that is equivalent to the $\eqso(\loge{1})$ formula $2$. 
First, if $\length{\bar{X}} = \length{\bar{x}} = 0$, then the function defined by $\alpha$ is never greater than 1. 
Therefore, suppose that $\length{\bar{X}} > 0$ or $\length{\bar{x}} > 0$, and consider some ordered structure $\A$. 
Since $\sem{\alpha}(\A) = 2$, there exist at least two assignments $(\bar{B}_1,\bar{b}_1,\bar{a}_1)$, $(\bar{B}_2,\bar{b}_2,\bar{a}_2)$ to $(\bar{X},\bar{x},\bar{y})$ such that for both, $\A\models\varphi(\bar{B}_i,\bar{b}_i,\bar{a}_i)$. Now consider the ordered structure $\A'$ that is obtained by concatenating $\A$ twice. Indeed, each half of $\A'$ is isomorphic to $\A$. Note that $\A'\models\varphi(\bar{B}_i,\bar{b}_i,\bar{a}_i)$ for $i = 1,2$ and there exists a third assignment $(\bar{B}_1',\bar{b}_1',\bar{a}_1')$ that is isomorphic to $(\bar{B}_1,\bar{b}_1,\bar{a}_1)$, in the other half of the structure, such that $\A'\models\varphi(\bar{B}_1',\bar{b}_1',\bar{a}_1')$. We have that $\sem{\alpha}(\A) \geq 3$ which leads to a contradiction.

@@

\sa{\bar{X}}\sa{\bar{x}}\sa{\bar{Y}}\sa{\bar{y}}[&(\varphi(\bar{X},\bar{x})\wedge\exists\bar{v}(\bar{v}_0 < \bar{x} \vee \bigvee_{i = 1}^{\length{\bar{X}}}X(\bar{v}_i))\wedge\bigwedge_{i = 1}^{\length{\bar{Y}}} \forall\bar{z}\neg Y_i(\bar{z}) \wedge \forall\bar{z}(\bar{y}\leq\bar{z}))\vee\\

@@

\sa{\bar{X}}&\sa{\bar{Y}}\sa{\bar{x}}\sa{\bar{y}}\sa{\bar{u}}[ \\
&(\varphi(\bar{X},\bar{x})\wedge(\bar{v}_0 < \bar{x} \vee \bigvee_{i = 1}^{\length{\bar{X}}}X(\bar{v}_i))\wedge \forall\bar{u}'((\bar{v}_0' < \bar{x} \vee \bigvee_{i = 1}^{\length{\bar{X}}}X(\bar{v}_i'))\to\bar{u}\leq\bar{u}') \wedge \\
&\bigwedge_{i = 1}^{\length{\bar{Y}}} \forall\bar{z}\neg Y_i(\bar{z}) \wedge \forall\bar{z}(\bar{y}\leq\bar{z}))\vee\\

@@

&(\bar{y} < \bar{w}_0 \vee \bigvee_{i = 1}^{\length{\bar{Y}}}\neg Y(\bar{w}_i))\wedge\forall\bar{u}'(\bar{y} < \bar{w}_0' \vee \bigvee_{i = 1}^{\length{\bar{Y}}}\neg Y(\bar{w}_i'))\to\bar{u}\leq\bar{u}')\vee \\
